\documentclass[11pt,a4paper]{scrartcl}
\usepackage[T1]{fontenc}
\usepackage{microtype}
\usepackage{lmodern}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{enumerate}
\usepackage{graphicx}
\usepackage{float}

\usepackage{listings}
\usepackage{color}
%% http://stackoverflow.com/questions/741985/latex-source-code-listing-like-in-professional-books
\usepackage{courier}
\definecolor{light-gray}{gray}{0.95}
\lstset{
  % language=C,
  basicstyle=\small\sffamily,
%   basicstyle=\small\ttfamily,
  numbers=left,
  numberstyle=\tiny,
  frame=tb,
%  columns=fullflexible,
%  showstringspaces=false,
	backgroundcolor=\color{light-gray},
	linewidth=\linewidth,       % Zeilenbreite
	breaklines=true,             % Zeileumbruch
	breakatwhitespace=false, %Umbruch an Leerzeichen
  tabsize=2,
  extendedchars=true,
  xleftmargin=17pt,
  framexleftmargin=17pt,
  abovecaptionskip=7pt,
%   frameround=tttt,
}

\def\ojoin{\setbox0=\hbox{$\bowtie$}%
  \rule[-.02ex]{.25em}{.4pt}\llap{\rule[\ht0]{.25em}{.4pt}}}
\def\leftouterjoin{\mathbin{\ojoin\mkern-5.8mu\bowtie}}
\def\rightouterjoin{\mathbin{\bowtie\mkern-5.8mu\ojoin}}
\def\fullouterjoin{\mathbin{\ojoin\mkern-5.8mu\bowtie\mkern-5.8mu\ojoin}}

\begin{document}

\author{Johannes Merkle\\Ralf Vogler}
\title{Query Optimization}
\subtitle{10. Exercise}

\maketitle

\section*{Exercise 1 - Bandwidth and seek time}
See folder \verb|Java| for the code (alternatively folder \verb|C#|). \verb:results_(Java|C#).txt: contains some test results. Drive \verb|C| is a SSD and drive \verb|G| is a NDAS connected over Gigabit-Ethernet. Results using other tools are shown in the images in this folder. The meassured seek time is probably slightly better because of the randomly generated seek positions (RNG based on time might yield the same value multiple times). The bandwith in the C\# version is off for small files since the Stopwatch-Class only uses milliseconds, wheras the Java version uses nanoseconds.

%\lstinputlisting[caption={Some test results},label=lst:results]{results.txt}


\section*{Exercise 2 - Break even point}
SSD: Assuming 170 MB/s bandwidth and 300 $\mu$s seek time:
\begin{align*}
t_{page} = 8KB / (170 MB/s) &= 45.9559 \mu s\\
s * t_{page} &\stackrel{!}{=} x * s * 300\mu s\\
x &= 0.1532 \approx 15\%
\end{align*}
Internal 7200rpm HDD: Assuming 80 MB/s bandwidth and 12 ms seek time:
\begin{align*}
t_{page} = 8KB / (80 MB/s) &= 97.6563 \mu s\\
s * t_{page} &\stackrel{!}{=} x * s * 12ms\\
x &= 0.0081 \approx 1\%
\end{align*}
where s is the total number of pages of the relation and x is the selectivity ratio.
If the query selects more than x\% of the relation, then it will be faster to do a scan instead of an index lookup.

\end{document}
